---
title: "Loading COMTRADE/WITS data"
author: "Eduard Bukin"
date: "August 23, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Ex

Load according to the following predefined parameters from two different sources (WITS and COMTRADE) and compare the results:

* Reporter: Poland
* Partner: All available partners
* Trade Flow: Import, Export
* Year: 2014
* Commodities: TOTAL  

Questions:  
1. Load both data sets to R under the names `ctPolandTotals` and `witsPolandTotals`  
2. Compare two data sets using different functions such as: `glimpse()`, `str()`, `nrows()`, `all_equal()` or more from `dplyr` package.
3. Briefly describe what is the difference between two data sets. 

## 1. Ex Answers

Step 1. general setup of the working area:
```{r }
library(tidyverse)
```

1. Loading CT data using the `read_csv()` because the are no customized function developed for reading regular CT data download. 0(Loading data after we manually loaded it from the website).

```{r pressure, echo=FALSE, warning=FALSE}
ctPolandTotals <- read_csv("01-data-import/ct_poland_data.csv")
glimpse(ctPolandTotals)
```

Loading WITS data using the `read_csv()`. (Loading data also after we manually loaded it from the website).

```{r}
witsPolandTotals <- read_csv("01-data-import/wits_poland_data.csv")
glimpse(witsPolandTotals)
```

It is clear, that both tables have dfferent number of rows as well as sets of columns. Therefore, we need to go one step ahead and use `dplyr` package and comands `select()` and `rename()` to clean and unify variables across two dataframes.

2. Cleaning WITS data frame.

```{r}
witsPolandTotals <- 
  witsPolandTotals %>% 
  select(Nomenclature, ReporterISO3, PartnerISO3, 
         ProductCode, Year, TradeFlowName, `TradeValue in 1000 USD`)
glimpse(witsPolandTotals)
```

Cleaning CT data frame.

```{r}
ctPolandTotals <- 
  ctPolandTotals %>% 
  select(Classification, `Reporter ISO`, `Partner ISO`, 
         `Commodity Code`, Year, `Trade Flow`, `Trade Value (US$)`) %>% 
  rename(Nomenclature = Classification, 
         ReporterISO3 = `Reporter ISO`, 
         PartnerISO3 = `Partner ISO`, 
         ProductCode = `Commodity Code`, 
         Year = Year, 
         TradeFlowName = `Trade Flow`, 
         `TradeValue in 1000 USD` = `Trade Value (US$)`)
glimpse(ctPolandTotals)
```

3. The differences are in the units of the `TradeValue` in both data sets, different number of rows, and codings for `ProductCode`.
