---
title: "Trade data sources overview" 
author: "Eduard Bukin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

To download entire report in a file, select the format: <a target="_blank" href="data-sources-compilation.pdf" class="btn btn-primary">PDF Report</a> or <a href="data-sources-compilation.docx" download="data-sources-compilation.docx" class="btn btn-primary">Word report</a>.

\newpage

# Trade classifications

```{r child = '00-classifications.Rmd'}
```

---

\pagebreak

# UN Comtrade data

```{r child = '00-comtrade.Rmd'}
```

---

\pagebreak

# World Bank data

```{r child = '00-worldbank.Rmd'}
```

---

\pagebreak

# World Trade Organization data

```{r child = '00-wto.Rmd'}
```

---

\pagebreak

# International Trade Center data

```{r child = '00-itc-trademap.Rmd'}
```

---

\pagebreak

# CEPII

```{r child = '00-cepi.Rmd'}
```

---

\pagebreak

# MIT-Harvard 'Economic Complexity' Solutions

```{r child = '00-economic-complexity.Rmd'}
```

---

\pagebreak

# Miscellaneous data sources

```{r child = '00-other-data-sources.Rmd'}
```

